Advances in Artificial Intelligence for Data Visualization: Automated Reading of Residual Plots with Computer Vision

### 2023-04-18 Milestone_presentation_2/Milestone_presentation_2/Milestone_presentation_2.Rmd
finish the intro
transfer file

---
title: "Automatic Visual Inference for Linear Regression"
subtitle: ""  
author: 
  - "Weihao (Patrick) Li"
date: '`r format(Sys.Date(), format="%B %d, %Y")`'
output:
  xaringan::moon_reader:
    seal: false
    css: [xaringan-themer.css, mine.css]
    nature:
      countdown: 55000
      beforeInit: "scripts/my_macro.js"
      countIncrementalSlides: false
      slideNumberFormat: "%current%/%total%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
---

@@

```{r}
library(RefManageR)
BibOptions(check.entries = FALSE,
           bib.style = "authoryear",
           cite.style = "apa",
           style = "Bibtex",
           # max.names = 3,
           longnamesfirst = FALSE,
           hyperlink = FALSE,
           dashed = FALSE)
myBib <- ReadBib("./Visual Inference.bib", check = FALSE)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
library(tidyverse)
library(visage)
style_solarized_light(
  background_color = "#FFFFFF",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono"),
  text_color = "#000000",
  text_font_size = "1.0rem",
  colors = c(myblue = "#006DAE", mywhite = "#FFFFFF")
)
```

@@

.pull-left-full[
<h1 class="myblue"> Automatic Visual Inference for Linear Regression </h1>

<h3 class="myblue"> First Milestone Presentation </h3>

<br>
<br>
<br>
<br>

<h3 class="myblue"> Weihao Li </h3>

<h3 class="myblue"> Supervised by Di Cook and Emi Tanaka </h3>

@@

.pull-left-center[
```{r}
library(visage)

eval_data <- readRDS("data/training_original_data_single.rds")

filter(eval_data, id == 9586, null == TRUE) %>%
  VI_MODEL$plot(theme = theme_light(base_size = 12), size = 1) +
  ylab("Residuals") +
  xlab("Fitted values")
# 68

```

.center[
.caption[
Fig. 1: *Residual plot of a simple linear regression.*
]
]
]

.pull-right[
# Visual Discoveries

@@

#### Can you find the evidence on heteroskedasticity?

@@

.pull-left-center[
```{r}
filter(eval_data, id == 9586, null == TRUE) %>%
  VI_MODEL$plot(theme = theme_light(base_size = 12), size = 1) +
  ylab("Residuals") +
  xlab("Fitted values")
# 68

```

.center[
.caption[
Fig. 1: *Residual plot of a simple linear regression.*
]
]
]

.pull-right[
# Visual Discoveries

#### Can you find the evidence on heteroskedasticity?

@@

.pull-left-center[
```{r}
filter(eval_data, id == 9586, null == TRUE) %>%
  VI_MODEL$plot(theme = theme_light(base_size = 12), size = 1) +
  ylab("Residuals") +
  xlab("Fitted values")
# 68

```

.center[
.caption[
Fig. 1: *Residual plot of a simple linear regression.*
]
]
]

.pull-right[
# Visual Discoveries

@@

#### Can you find the evidence on heteroskedasticity?

#### However, the residuals are actually obtained from a **correctly specified classical normal linear model**!

@@

.pull-left-center[
```{r}
filter(eval_data, id == 9586, null == TRUE) %>%
  VI_MODEL$plot(theme = theme_light(base_size = 12), size = 1) +
  ylab("Residuals") +
  xlab("Fitted values")
# 68

```

.center[
.caption[
Fig. 1: *Residual plot of a simple linear regression.*
]
]
]

.pull-right[
# Visual Discoveries

@@

#### Can you find the evidence on heteroskedasticity?

@@

#### However, the residuals are actually obtained from a **correctly specified classical normal linear model**!

#### Unsecured and unconfirmed visual discoveries will lead to **over or under-interpretations of the data**.

@@

.pull-left[
# Visual Inference & Lineup Protocol

@@

#### **Visual inference** was introduced by `r BibOptions(max.names = 1);Citet(myBib, "buja_statistical_2009")` as an inferential framework to extend confirmatory statistics to visual discoveries.

@@

.pull-right-center[
```{r}
mod <- HETER_MODEL$fit(filter(eval_data, id == 9586, null == FALSE))

manual_lineup <- data.frame()

for (i in 1:19) {
  tmp_dat <- HETER_MODEL$null_resid(filter(eval_data, id == 9586, null == FALSE), mod)
  tmp_dat$k <- i
  manual_lineup <- bind_rows(manual_lineup, tmp_dat)
}

manual_lineup <- bind_rows(manual_lineup, mutate(filter(eval_data, id == 9586, null == TRUE), k = 20))

VI_MODEL$plot_lineup(manual_lineup, 
                     theme = theme_light(),
                     remove_grid_line = TRUE,
                     size = 0.3
                     ) +
  xlab("Fitted values") +
  ylab("Residuals")
```

.center[
.caption[
Fig. 2: _A lineup of 20 residual plots of a simple linear regression. (Plot No.20 is Fig.1) **Can you find the most different one?**_ 
]
]

@@

#### **Visual inference** was introduced by `r BibOptions(max.names = 1);Citet(myBib, "buja_statistical_2009")` as an inferential framework to extend confirmatory statistics to visual discoveries.

#### A **lineup** consists of $m$ randomly placed plots, where one plot is the **actual data plot** and the remaining $m-1$ plots (**null plots**) contain data consistent with the null hypothesis.

@@

.pull-right-center[
```{r}
VI_MODEL$plot_lineup(manual_lineup, 
                     theme = theme_light(),
                     remove_grid_line = TRUE,
                     size = 0.3
                     ) +
  xlab("Fitted values") +
  ylab("Residuals")
```

.center[
.caption[
Fig. 2: _A lineup of 20 residual plots of a simple linear regression. (Plot No.20 is Fig.1) **Can you find the most different one?**_ 
]
]

@@

.pull-left[
# Visual Inference & Lineup Protocol

@@

#### **Visual inference** was introduced by `r BibOptions(max.names = 1);Citet(myBib, "buja_statistical_2009")` as an inferential framework to extend confirmatory statistics to visual discoveries.

#### A **lineup** consists of $m$ randomly placed plots, where one plot is the **actual data plot** and the remaining $m-1$ plots (**null plots**) contain data consistent with the null hypothesis.

#### To perform a **visual test**, observer(s) will be asked to select the **most different plot(s)** from the lineup.

@@

.pull-right-center[
```{r}
VI_MODEL$plot_lineup(manual_lineup, 
                     theme = theme_light(),
                     remove_grid_line = TRUE,
                     size = 0.3
                     ) +
  xlab("Fitted values") +
  ylab("Residuals")
```

.center[
.caption[
Fig. 2: _A lineup of 20 residual plots of a simple linear regression. (Plot No.20 is Fig.1) **Can you find the most different one?**_ 
]
]

@@

# $p$-value of Lineup Protocol

#### Under the null hypothesis, it is expected that the actual data plot would have **no distinguishable difference** with the null plots.

# $p$-value of Lineup Protocol

#### Under the null hypothesis, it is expected that the actual data plot would have **no distinguishable difference** with the null plots.

#### Suppose observers are allowed to select **only one plot**. The probability of the observer correctly picks the actual data plot is $1/m$.

# $p$-value of Lineup Protocol

#### Under the null hypothesis, it is expected that the actual data plot would have **no distinguishable difference** with the null plots.

#### Suppose observers are allowed to select **only one plot**. The probability of the observer correctly picks the actual data plot is $1/m$.

#### If we involve $K$ independent observers in a visual test, and let $X$ be a random variable denoting the number of observers correctly picking the actual data plot. Under the null hypothesis $X ∼ Binom(K, 1/m)$.

# $p$-value of Lineup Protocol

#### Under the null hypothesis, it is expected that the actual data plot would have **no distinguishable difference** with the null plots.

#### Suppose observers are allowed to select **only one plot**. The probability of the observer correctly picks the actual data plot is $1/m$.

#### If we involve $K$ independent observers in a visual test, and let $X$ be a random variable denoting the number of observers correctly picking the actual data plot. Under the null hypothesis $X ∼ Binom(K, 1/m)$.

#### The p-value is given as $$P(X \geq x) = \sum_{i=x}^{K}{{K}\choose{i}}\left(\frac{1}{m}\right)^i\left(\frac{m-1}{m}\right)^{k-i}.$$

# $p$-value of Lineup Protocol

#### Under the null hypothesis, it is expected that the actual data plot would have **no distinguishable difference** with the null plots.

#### Suppose observers are allowed to select **only one plot**. The probability of the observer correctly picks the actual data plot is $1/m$.

#### If we involve $K$ independent observers in a visual test, and let $X$ be a random variable denoting the number of observers correctly picking the actual data plot. Under the null hypothesis $X ∼ Binom(K, 1/m)$.

#### The p-value is given as $$P(X \geq x) = \sum_{i=x}^{K}{{K}\choose{i}}\left(\frac{1}{m}\right)^i\left(\frac{m-1}{m}\right)^{k-i}.$$

#### The null hypothesis will be rejected if $P(X \geq x) < \alpha$, the desired significance level.

![:scale 80%](images/Powerloom_weaving_in_1835.jpg)

![:scale 80%](images/Maquina_vapor_Watt_ETSIIM.jpg)

.caption[
Fig. 3: _A Roberts loom in a weaving shed in 1835 and a Watt steam engine. (https://en.wikipedia.org/wiki/Industrial_Revolution)_ 
]

@@

.pull-right-120[
# Limitation of Lineup Protocol

#### 1. Human can not evaluate lineup consisted of a large number of plots.
#### 2. Human can not evaluate a large number of lineups.
#### 3. Lineup protocol is unfriendly to vision-impaired people.
#### 4. Evaluation of lineup is high in labour cost and time consuming.

@@

![:scale 80%](images/Powerloom_weaving_in_1835.jpg)

![:scale 80%](images/Maquina_vapor_Watt_ETSIIM.jpg)

.caption[
Fig. 3: _A Roberts loom in a weaving shed in 1835 and a Watt steam engine._ 
]

.pull-right-120[
# Limitation of Lineup Protocol

@@

#### 1. Human can not evaluate lineup consisted of a large number of plots.
#### 2. Human can not evaluate a large number of lineups.
#### 3. Lineup protocol is unfriendly to vision-impaired people.
#### 4. Evaluation of lineup is high in labour cost and time consuming.

### **We need a "steam engine" for visual test!**

@@

# Computer Vision Model

#### Large-scale evaluation of visual tests is not possible without the use of technology and machines.

#### Modern **computer vision model** could be a promising solution to this problem. It is usually built on a **deep neural network** called **convolutional neural network** (CNN) `r BibOptions(max.names = 1);Citep(myBib, "fukushima_neocognitron_1982")`.

![](images/Typical_cnn.png)

.caption[
Fig. 4: _A typical CNN with convolutional layers, pooling layers and fully-connected layers. (https://en.wikipedia.org/wiki/Convolutional_neural_network)_ 
]

exclude: true

# Computer Vision Model - Convolutional Layer

#### **Convolutional layers** take advantages of the hierarchical pattern in data and provide regularized versions of fully-connected layers. It downscales and transforms the image by summarising information in a small space.

.center[![:scale 70%](images/conlayer.png)]

.center[.caption[
Fig. 5: _Illustration of convolution operation with two filters._ 
]]

exclude: true

# Computer Vision Model - Pooling Layer

#### **Pooling layers** apply non-linear down-sampling on the feature maps. This helps reducing the number of parameters in the neural network so that the problem of overfiting can be controlled.

#### One of the most popular pooling operations is **max pooling**. It divides the feature map into several rectangle regions and takes the maximum to form a new feature map.
]

.pull-right-center[
![](images/Max_pooling.png)

.caption[
Fig. 6: _Illustration of max pooling operation._ 
]
]

exclude: true

# Computer Vision Model - Fully-connected Layer

.pull-left-center-80[
![:scale 50%](images/fnn.png)

.caption[
Fig. 7: _Illustration of a fully-connected layer._ 
]
]

#### **Fully-connected layer** is building block of neural network. It peformes matrix multiplication on the inputs and the weights.

#### This layer is primarily used to process information presented in the feature maps before making the prediction. The output of the layer will be passed to a special layer called **loss layer** for computing the loss or error of the prediction.

# Simulation Setup - Cubic Model

#### We simulated data from two linear regression model with **non-linearity** and **heteroskedasticity** issues

#### The first model is a **cubic model** which can be expressed by: $$\boldsymbol{Y}= 1 + (2-c)\boldsymbol{X} + c\boldsymbol{Z} + a[(2-c)\boldsymbol{X}]^2+a(c\boldsymbol{Z})^2+b[(2-c)\boldsymbol{X}]^3+b(c\boldsymbol{Z})^3+\boldsymbol{\varepsilon},$$ where $c \in (0,2)$, $a \in (-3,3)$, $b \in (-3,3)$, $\boldsymbol{\varepsilon}\overset{iid}{\sim} N(\boldsymbol{0},\sigma^2\boldsymbol{I})$, $\boldsymbol{Y}$, $\boldsymbol{X}$ and $\boldsymbol{Z}$ are $n\times1$ matrices.

.pull-left[
#### $X_i$, $i=1,...,n$ follows one of the four distributions:
- $N(0, 0.09)$
- $U(-1, 1)$
- $Lognormal(0,0.36)/3 - 1$
- $-Lognormal(0,0.36)/3 + 1$

#### Most the realizations will fall between $-1$ and $1$.
]

.pull-right[
#### $Z_i$, $i=1,...,n$ follows one of the two distributions:
- $U(-1, 1)$
- Discrete uniform distribution with $z_n$ outcomes ranged from $-1$ to $1$

#### $z_n$ is sampled from $\{z_n: 9<z_n<21 \text{ and } z_n \in \mathbb{Z} \}$.
]

<!-- #### The distribution of $Z_i$, $i=1,...,n$, has $50$% chance to be a uniform distribution ranged from $-1$ to $1$, and $50$% chance to be a discrete uniform distribution with $z_n$ outcomes simulated from a uniform distribution ranged from $-1$ to $1$. $z_n$ itself is a discrete uniform distribution with outcomes $\{x: 9<x<21 \text{ and } x \in \mathbb{Z} \}$, which defines the number of possible values $Z_i$ could take. -->

# Simulation Setup - Cubic Model

#### The simulated data is fitted by OLS with equation: $$\boldsymbol{Y}= \beta_0 + \beta_1\boldsymbol{X} + \beta_2\boldsymbol{Z} +\boldsymbol{u},$$ where $\boldsymbol{u} \sim N(\boldsymbol{0}, \sigma^2_u\boldsymbol{I})$.

#### Let $\boldsymbol{X}_a=[\boldsymbol{1},\boldsymbol{X},\boldsymbol{Z}]$ denotes the set of regressors in matrix form. Then, the **residuals** obtained from the null model are $$\boldsymbol{e} \sim N(\boldsymbol{R}_a\boldsymbol{X}_b\boldsymbol{\beta}_b, \sigma^2\boldsymbol{R}_a),$$ where $\boldsymbol{R}_a=\boldsymbol{I}-\boldsymbol{X}_a(\boldsymbol{X}_a'\boldsymbol{X}_a)^{-1}\boldsymbol{X}_a'$, $\boldsymbol{X}_b=[\boldsymbol{X}^2,\boldsymbol{Z}^2,\boldsymbol{X}^3,\boldsymbol{Z}^3]$ and $\boldsymbol{\beta}_b=(a(2-c)^2,ac^2,b(2-c)^3,bc^3)'$.

#### Under the null hypothesis, $\boldsymbol{e} \sim N(\boldsymbol{0}, \sigma^2\boldsymbol{R}_a)$. The **effect size** is derived as: $$ES=n^{-1}||\sigma^{-1}\boldsymbol{R}_a^{-\frac{1}{2}}\boldsymbol{R}_a\boldsymbol{X}_b\boldsymbol{\beta}_b||^2=n^{-1}\sigma^{-2}||\boldsymbol{R}_a^{\frac{1}{2}}\boldsymbol{X}_b\boldsymbol{\beta}_b||^2\approx n^{-1}\sigma^{-2}||diag(\boldsymbol{R}_a)^{\frac{1}{2}}\boldsymbol{X}_b\boldsymbol{\beta}_b||^2,$$ where $diag(\boldsymbol{R}_a)$ is the diagonal matrix constructed from the diagonal elements of $\boldsymbol{R}_a$.

.pull-left[
```{r}
mod <- cubic_model(a = 5, b = 10, c = 1)
mod$plot_lineup(mod$gen_lineup(300), remove_grid_line = TRUE, theme = theme_light()) + xlab("Fitted values") + ylab("Residuals")
```

.caption[
Fig. 5: _A lineup of residual plots generated with the cubic model._ 
]
]

.pull-right[
```{r}
mod <- cubic_model(-1, -1, 1, 0.5)
dat <- mod$gen(500, fit_model = TRUE)
mod$plot(dat, theme = theme_light(base_size = 10), size = 1) +
  ggtitle(glue::glue("Effect size: {round(mod$effect_size(dat), 2)}")) + xlab("Fitted values") + ylab("Residuals") -> p1

mod <- cubic_model(-2, -1, 1, 0.5)
dat <- mod$gen(500, fit_model = TRUE)
mod$plot(dat, theme = theme_light(base_size = 10), size = 1) +
  ggtitle(glue::glue("Effect size: {round(mod$effect_size(dat), 2)}")) + xlab("Fitted values") + ylab("Residuals") -> p2

mod <- cubic_model(-5, -5, 1, 0.5)
dat <- mod$gen(500, fit_model = TRUE)
mod$plot(dat, theme = theme_light(base_size = 10), size = 1) +
  ggtitle(glue::glue("Effect size: {round(mod$effect_size(dat), 2)}")) + xlab("Fitted values") + ylab("Residuals") -> p3

mod <- cubic_model(-20, -20, 1, 0.5)
dat <- mod$gen(500, fit_model = TRUE)
mod$plot(dat, theme = theme_light(base_size = 10), size = 1) +
  ggtitle(glue::glue("Effect size: {round(mod$effect_size(dat), 2)}")) + xlab("Fitted values") + ylab("Residuals") -> p4

list(p1, p2, p3, p4) %>% patchwork::wrap_plots(ncol=2)
```

.caption[
Fig. 6: _Residual plots of cubic model under different effect sizes._ 
]
]

# Simulation Setup - Heteroskedasticity Model

#### The second model is a heteroskedasticy model which can be expressed by: $$Y_i = 1 + X_i + \varepsilon_i, ~i = 1,...,n,$$ where $a \in \{-1,0,1\}$, $b\in (0,32)$ and $\varepsilon_i \overset{iid}{\sim} N(0,1+b(2-|a|)(X_i-a)^2)$.

>>>>

---
title: "Advances in Artificial Intelligence for Data Visualization: Automated Reading of Residual Plots with Computer Vision"
subtitle: ""  
author: 
  - "Weihao (Patrick) Li"
date: '`r format(Sys.Date(), format="%B %d, %Y")`'
output:
  xaringan::moon_reader:
    seal: false
    css: [xaringan-themer.css, mine.css]
    nature:
      countdown: 55000
      beforeInit: "scripts/my_macro.js"
      countIncrementalSlides: false
      slideNumberFormat: "<%current%>"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
---

@@

```{r}
set.seed(10086)
library(RefManageR)
BibOptions(check.entries = FALSE,
           bib.style = "authoryear",
           cite.style = "apa",
           style = "Bibtex",
           # max.names = 3,
           longnamesfirst = FALSE,
           hyperlink = FALSE,
           dashed = FALSE)
myBib <- ReadBib("./Visual Inference.bib", check = FALSE)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
library(tidyverse)
library(visage)
style_solarized_light(
  background_color = "#FFFFFF",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i", "400i", "600"),
  code_font_google   = google_font("Fira Mono"),
  text_color = "#000000",
  text_font_size = "1.0rem",
  header_h1_font_size = "2.0rem",
  colors = c(myblue = "#006DAE", mywhite = "#FFFFFF")
)
```

@@

.pull-left-full[
<h2 class="myblue"> Advances in Artificial Intelligence for Data Visualization: Automated Reading of Residual Plots with Computer Vision </h2>

<h3 class="myblue"> Progress Review Presentation </h3>

<h4 class="myblue"> Weihao (Patrick) Li </h3>

<h4 class="myblue"> Supervised by Di Cook, Emi Tanaka and Susan VanderPlas </h3>

@@

.center[
# 📚Thesis Structure📚
]

#### 1. **Exploring the application of visual inference in regression diagnostics and comparing it with conventional hypothesis tests.**

#### 2. Designing an automated visual inference system to assess lineups of residual plots of classical normal linear regression model.

#### 3. Deploying the automatic visual inference system as an online application and publishing the relevant open-source software.

@@

This presentation will be focused on the **first project**.

# A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

#### Weihao Li, Dianne Cook, Emi Tanaka and Susan VanderPlas (2023)

.center[

# 🔍Regression Diagnostics🔍

**Diagnostics** are the key to determining whether there is anything **importantly wrong** with a model.

**Residuals** summarise what is **not captured by the regression model**.

**Residual plots** are commonly used to diagnose **non-linearity** and **heteroskedasticity**. **Non-normality** is usually harder to detect from a residual plot. A favourable graphical summary for this task is the **quantile-quantile plot**.

.image-top-bottom-0[

```{r fig.height = 2.5, fig.width = 10}
set.seed(10086)

# Base model
mod <- poly_model(include_z = FALSE, sigma = 0.25)

# Base data
dat <- mod$gen(300)

bind_rows(
  dat %>%
    mutate(type = "(A) Good residuals"),
  mod$set_prm("include_z", TRUE)$
    gen(300, computed = select(dat, x, e)) %>%
    mutate(type = "(B) Non-linearity"),
  heter_model(b = 64)$
    gen(300, computed = select(dat, x, e)) %>%
    mutate(type = "(C) Heteroskedasticity"),
  heter_model(b = 0, e = rand_lognormal(sigma = 0.5))$
    gen(300, computed = select(dat, x)) %>%
    mutate(type = "(D) Non-normality")
) %>%
  mod$plot(remove_grid_line = TRUE, theme = theme_light(base_size = 12)) +
  facet_wrap(~type, ncol = 4, scales = "free") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank()) +
  xlab("Fitted values") +
  ylab("Residuals")
```

.center[
.caption[
Fig. 1: _Example residual plots: (A) classically good looking residuals, (B) non-linear pattern indicates that the model has not captured a non-linear association, (C) heteroskedasticity indicating that variance around the fitted model is not uniform, and (D) non-normality where the residual distribution is not symmetric around 0._
]
]

.center[
# 🔍Regression Diagnostics🔍
]

Many different **residual-based hypothesis tests** are available to detect specific model defects.

.pull-left-33p[

Non-linearity:

- **F-test**

- **RESET test** (Ramsey 1969)

.pull-left-33p[

Heteroskedasticity:

- **White test** (White 1980)

- **BP test** (Breusch and Pagan 1979)

.pull-left-33p[

Non-normality:

- **SW test** (Shapiro and Wilk 1965)

- **Jarque–Bera test** (Jarque and Bera 1980)

.float-left[
But there are **drawbacks** of using them.

- Most residual-based tests for particular types of departure are also sensitive to **other types of departures**, resulting in a phenomenon known as the "**Type III error**" (Cook and Weisberg 1982).

- Other problems such as **outliers** can trigger the rejection (Cook and Weisberg 1999).

@@

.image-top-bottom-0[

```{r}
set.seed(10086)
# Base model
mod <- poly_model(include_z = FALSE, sigma = 0.25)
# Base data
dat <- mod$gen(300)
bind_rows(
  dat %>%
    mutate(type = "(A) Good residuals"),
  mod$set_prm("include_z", TRUE)$
    gen(300, computed = select(dat, x, e)) %>%
    mutate(type = "(B) Non-linearity"),
  heter_model(b = 64)$
    gen(300, computed = select(dat, x, e)) %>%
    mutate(type = "(C) Heteroskedasticity"),
  heter_model(b = 0, e = rand_lognormal(sigma = 0.5))$
    gen(300, computed = select(dat, x)) %>%
    mutate(type = "(D) Non-normality")
) %>%
  mod$plot(remove_grid_line = TRUE, theme = theme_light(base_size = 12)) +
  facet_wrap(~type, ncol = 2, scales = "free") +
  theme(axis.text = element_blank()) +
  xlab("Fitted values") +
  ylab("Residuals")
```

.center[
.caption[
Fig. 1: _Example residual plots: (A) classically good looking residuals, (B) non-linear pattern indicates that the model has not captured a non-linear association, (C) heteroskedasticity indicating that variance around the fitted model is not uniform, and (D) non-normality where the residual distribution is not symmetric around 0._
]
]

.pull-right-120[

# 🔍Regression Diagnostics🔍

BP and SW tests reject the residual plots exhibiting structure that **they weren’t designed for** (**type III error**).

```{r}
set.seed(10086)
# Base model
mod <- poly_model(include_z = FALSE, sigma = 0.25)
# Base data
dat0 <- mod$gen(300)
# Replicate data in Figure 1
dat1 <- mod$set_prm("include_z", TRUE)$gen(300, computed = select(dat0, x, e))
dat2 <- heter_model(b = 64)$gen(300, computed = select(dat0, x, e))
dat3 <- heter_model(b = 0, e = rand_lognormal(sigma = 0.5))$
               gen(300, computed = select(dat0, x))
dat_list <- list(dat0, dat1, dat2, dat3)
table_dat <- data.frame(plot = c("A", "B", "C", "D"), 
                        model = c("None", 
                                  "Non-linearity", 
                                  "Heteroskedasticity", 
                                  "Non-normality"),
                        r = map_dbl(dat_list, 
                                    ~POLY_MODEL$test(.x, 
                                                     test = "RESET", 
                                                     power = 2:4)$p_value),
                        b = map_dbl(dat_list, 
                                    ~HETER_MODEL$test(.x)$p_value),

                        # SW test has not been included by visage
                        s = map_dbl(dat_list, 
                                    ~shapiro.test(.x$.resid)$p.value)) %>%

  # 3 decimal points
  mutate(across(r:s, ~ format(round(.x, digits = 3), nsmall = 3))) %>%

  # Italic if reject
  mutate(across(r:s, ~ kableExtra::cell_spec(.x, italic = as.numeric(.x) <= 0.05, color = ifelse(as.numeric(.x) <= 0.05, "red", "black"))))
  table_dat %>%
  knitr::kable(col.names = c("Plot", 
                             "Departures", 
                             "RESET", 
                             "BP", 
                             "SW"),
               align = "llrrr",
               escape = FALSE,
               format = "html",
               booktabs = TRUE)
```

.center[
.caption[
Tab. 1: _Statistical significance testing for departures from good residuals. Shown are the \\(p\\)-values calculated for the RESET, the BP and the SW tests._
]
]

@@

.pull-right-center[

.image-top-bottom-0[

```{r}
set.seed(10086)

# Base model
mod <- poly_model(sigma = 8)
dat <- mod$gen(1000)
this_p_value <- mod$test(dat, test = "RESET", power = 2:4)
dat %>%
mod$plot(remove_grid_line = TRUE, theme = theme_light(base_size = 12)) +
  theme(axis.text = element_blank()) +
  xlab("Fitted values") +
  ylab("Residuals")
```

.center[
.caption[
Fig. 2: _One classically good looking residual plot but rejected by the RESET test with $p\text{-value} = `r format(round(this_p_value$p_value, digits = 3), nsmall = 3)`$. The model is misspecified but the departure from the model assumption is small._
]
]

# 🔍Regression Diagnostics🔍

@@

With **large sample sizes**, hypothesis tests may reject a residual plot when there is only a **slight departure** (Belsley, Kuh, and Welsch 1980).

<!-- - While such rejections may be **statistically correct**, their **sensitivity** may render the results **impractical**. **Minor defects** in the model are **unlikely** to have a **material impact** and may be best disregarded for practical purposes.  -->

Residual plot diagnostics:
  - **Identify** potential issues that could lead to **incorrect conclusions or errors in subsequent analyses**.

@@

.center[
# 📜Liteature of Regression Diagnostics📜
]

### Grahpical approaches (plots) are the recommended methods for diagnosing model fits.

- Draper and Smith (1998) and Belsley, Kuh, and Welsch (1980):
> Residual plots are usually revealing when the assumptions are violated.

- Cook and Weisberg (1982):
> Formal tests and graphical procedures are complementary and both have a place in residual analysis, but graphical methods are easier to use.

- Montgomery and Peck (1982):
> Residual plots are more informative in most practical situations than the corresponding conventional hypothesis tests.

.image-top-bottom-0[

```{r}
set.seed(10131)
mod <- heter_model(b = 0, x = rand_lognormal())
ori_dat <- mod$gen(300)

ori_dat %>%
  VI_MODEL$plot(theme = theme_light(base_size = 12)) +
  xlab("Fitted values") +
  ylab("Residuals")
```

.center[
.caption[
Fig. 3: _Residual plot of a correctly specified model exhibiting a heteroskedasticity pattern (triangle shape)._
]
]

# 🤔False Visual Discoveries🤔

@@

Can you find the evidence on heteroskedasticity?

@@

However, the residuals are actually obtained from a **correctly specified classical normal linear model**!

**Unconfirmed** visual discoveries could result in **over or under-interpretations of the data**.

@@

# 🔬Visual Inference🔬

@@

If we embed the residual plot in **a matrix of plots containing residuals simulated from the fitted model**, we will find that it is **not uncommon** for residual plots to exhibit a triangle shape.

- Can you find the **most different plot**?

- Is it the residual plot presented at the previous page?

- If it is not, then the visual pattern is **over-interpreted**!

### The visual discovery is calibrated via comparison.

@@

.pull-right-center[

```{r}
set.seed(10131)
mod$gen_lineup(300, k = 20, pos = 8) %>%
  filter(null != FALSE) %>%
  bind_rows(ori_dat %>% mutate(k = 8, null = FALSE)) %>%
  VI_MODEL$plot_lineup(theme = theme_light(base_size = 12),
                       remove_grid_line = TRUE,
                       remove_axis = TRUE)
```
.center[
.caption[
Fig. 4: _A lineup of 20 residual plots of a simple linear regression. (Plot No.8 is Fig.3) Can you find the most different one?_
]
]

@@

This procedure is called **visual inference**. It was introduced by Buja, et al. (2009) as an inferential framework to extend confirmatory statistics to visual discoveries.

A **lineup** consists of $m$ randomly placed plots, where one plot is the **data plot** and the remaining $m − 1$ plots (**null plots**) contain data consistent with the null hypothesis.

To perform a **visual test**, observer(s) will be asked to select the **most different plot(s)** from the lineup.

@@

.pull-right-center[

```{r}
set.seed(10131)
mod$gen_lineup(300, k = 20, pos = 8) %>%
  filter(null != FALSE) %>%
  bind_rows(ori_dat %>% mutate(k = 8, null = FALSE)) %>%
  VI_MODEL$plot_lineup(theme = theme_light(base_size = 12),
                       remove_grid_line = TRUE,
                       remove_axis = TRUE)
```
.center[
.caption[
Fig. 4: _A lineup of 20 residual plots of a simple linear regression. (Plot No.8 is Fig.3) Can you find the most different one?_
]
]

@@

.center[
# 🎲Simulate Residuals from the Assumed Model🎲
]

Data used in the $m − 1$ null plots needs to be simulated.

For classical normal linear regression model, the **residual rotation** technique (Buja, et al. 2009) can be applied:

<p> 1. Generate independent random draws \(w_i, i=1,...,n\) from \(N(0, 1)\). </p>

<p> 2. Regress \(w\) on \(x\) and obtain the residuals \(r_i, i=1,...,n\). </p>

<p> 3. Rescale \(r_i\) by \(\sqrt{RSS_{old}/RSS_{new}}\) to obtain the simulated residuals, where \(RSS_{old}\) is the residual sum of square of the original regression, and \(RSS_{new}\) is the residual sum of square of the regression fitted on step 2. </p>

.center[
# 📈Statistical Significance📈
]

@@

Under the null hypothesis, it is expected that the data plot would have **no distinguishable difference** with the null plots.

Suppose observers are allowed to **select only one plot**. The probability of the observer correctly picks the actual data plot is $1/m$.

However, if we involve $K$ **independent observers** in a visual test, and show them the **same lineup**, there will be possible **dependencies** in the visual test due to **repeated evaluations** of the same lineup.

```{r}
alpha_sample <- rgamma(10, 1, 1)
data.frame(x = 0, y = 0, k = 1:20, alpha = alpha_sample/sum(alpha_sample)) %>%
ggplot() +
  geom_text(aes(x, y, label = paste0("theta[", k, "]~'='~", round(alpha, 2))), parse = TRUE, size = 5) +
  facet_wrap(~k) +
  theme_light(base_size = 12) +
  theme(panel.grid = element_blank(),
        axis.line = element_blank(),
        axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())
```

.center[
.caption[
Fig. 5: _Example of the probability of each plot being selected \\(\theta_i\\) under the null hypothesis with parameter \\(\alpha = 1\\)._
]
]

@@

# 📈Statistical Significance📈

VanderPlas et al. (2021) addresses this by modelling the probability of a plot $i$ being selected from a lineup as $\theta_i$:

$$\theta_i \sim Dirichlet(\alpha) \text{ for } i=1,...,m \text{ and } \alpha > 0.$$

- This assumes the **attractiveness distribution** of each plot.

For **small** $\alpha$:
- Only **a few plots** are attractive and tend to be selected.

For **large** $\alpha$:
- The distribution of the probability of each plot being selected is more **even**.

@@

```{r}
data.frame(k = 1:20, 
           ci = c(1, 0, 0.25, 0, 0, 0.2, 0.45, 0, 5.95, 0, 0, 0, 0, 1.7, 0, 0.2, 0, 1, 0, 0.25)) %>%
ggplot() +
  geom_text(aes(0, 0, label = paste0("c[", k, "]~'='~", ci)), parse = TRUE, size = 5) +
  facet_wrap(~k) +
  theme_light(base_size = 12) +
  theme(panel.grid = element_blank(),
        axis.line = element_blank(),
        axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())
```

.center[
.caption[
Fig. 5: _Example of the number of times a plot being selected \\(c_i\\). The sum of all \\(c_i\\) is 11 because the lineup is evaluated by 11 subjects._
]
]

@@

# 📈Statistical Significance📈

The **number of times** plot $i$ being selected in $K$ evaluations is denoted as $c_i$.

- In case subject $j$ makes **multiple selections**, $1/s_j$ will be added to $c_i$ instead of one, where $s_j$ is the **number of plots subject $j$ selected** for $j=1,...K$. 
- This ensures $\sum_{i}c_i=K$.

@@

.center[
# 📈Statistical Significance📈
]

@@

Since we are only interested in the **selections of the data plot $i$**, the **marginal model** can be simplified to a **beta-binomial model**. VanderPlas et al. (2021) derives the visual $p$-value as

$$P(C \geq c_i) = \sum_{x=c_i}^{K}{K \choose x}\frac{B(x + \alpha, K - x + (m - 1)\alpha)}{B(\alpha, (m-1)\alpha)},\text{ for } c_i \in \mathbb{Z}_0^+$$

where $B(.)$ is the beta function defined as

$$B(a, b) = \int_{0}^{1}t^{\alpha - 1}(1-t)^{b-1}dt,\text{ where } a,b>0.$$

We extend the equation to **non-negative real number $c_i$** by applying a linear approximation

$$
P(C \geq c_i) = P(C \geq \lceil c_i \rceil) + (\lceil c_i \rceil - c_i) P(C = \lfloor c_i \rfloor), \text{ for } c_i \in \mathbb{R}_0^+.
$$

@@

.center[
# 💪Power of Visual Tests💪
]
